Beyond conceptual eco-physiological hypotheses: insights from a general size and trait-based model of ectotherm thermal performance

### 2018-03-20 draft_model_only.Rnw
add theory discussion in intro

Increasing temperatures under climate change are thought to affect individual physiology of ectotherms through increases in metabolic demands, leading to changes in species performance with concomitant effects on species ecology. Although intuitively appealing, the driving mechanisms behind thermal performance is contested: thermal performance (e.g., growth) appears correlated with metabolic scope (i.e., oxygen availability for activity) for a number of species, but a substantial number of datasets do not support oxygen limitation of long-term performance. Whether or not oxygen limitations via the metabolic scope, or a lack thereof, have major ecological consequences remains a highly contested question. Here, we propose a general size and trait-based model of energy and oxygen budgets to determine the relative influence of thermal sensitivity of metabolic rates, oxygen limitation, and environmental conditions on thermal performance. We show that oxygen limitation is not necessary to explain performance curves that are correlated with metabolic scope. Oxygen can drastically limit performance and fitness at temperature extremes, but changes in thermal performance are primarily driven by the interplay of changing metabolic rates and species ecology. Furthermore, our model reveals that fitness trends with temperature likely oppose trends in growth and efficiency, suggesting a potential explanation for the paradox that species often occur at lower temperatures than their growth-optimum. We suggest that this general model provides a mechanistic underpinning that can provide general and realistic predictions about temperature impacts on ectotherm performance and effects of climate change on species with different metabolic and ecological traits.

@@

There are few general concepts and models that can explain thermal performance and ecological responses to temperature. To our knowledge, no general theoretical framework exists to quantitatively explain and predict changes in fundamental ecological rates, such as observed changes of attack rates with temperature (\cite{englund_temperature_2011,rall_universal_2012}), from fundamental physiological processes. Instead, and as advocated by the metabolic theory of ecology (\cite{brown_toward_2004}), ecological theory often treats ecological rates as being directly temperature dependant, without a direct link to the underlying physiological drivers (\cite[e.g., ][]{vucic-pestic_warming_2011, guiet_effects_2016}).

Although ecological rates seem to follow some general patterns in the response to temprature (\cite{englund_temperature_2011,rall_universal_2012}), there is also significant heterogeneity between species and trait groups. A phenomenological description that assumes a general ecological temperature response thus leads to difficulties with extrapolation when observed ecological responses are heterogeneous across species or other model components (\cite[e.g., fish in marine ecosystem models; ][]{guiet_effects_2016}). In this case, a deeper understanding of the underlying drivers of thermal responses may be necessary in order to derive general predictions about ecological responses to changing temperatures (\cite[e.g., ][]{vucic-pestic_warming_2011}). Since the primary effect of temperature of organisms is on individual physiology, a general model to explain ecological response should be grounded in physiology.

@@

This view was encapsulated in the Theory of Oxygen and Capacity Limitation of Temperature (\cite{portner_oxygen-and_2010}), which suggests that the decrease in metabolic scope towards extreme temperatures limits species ability to sustain core functions such as foraging and growth (i.e., functions beyond SM). In some species, however, maximum metabolic rate increases steadily (\cite{lefevre_are_2016, verberk_does_2016}), suggesting that oxygen may not be the limiting factor at high temperatures. Indeed, it has been argued that oxygen is unlikely to determine performance for most species over most of their temperature range as oxygen limits are rarely reached during normal activity (\cite{jutfelt_oxygen-and_2018, holt_climate_2015}).

>>>>

Increasing temperatures under climate change are thought to affect individual physiology of ectotherms through increases in metabolic demands, leading to changes in species performance with concomitant effects on species ecology. Although intuitively appealing, the driving mechanisms behind thermal performance is contested: thermal performance (e.g., growth) appears correlated with metabolic scope (i.e., oxygen availability for activity) for a number of species, but a substantial number of datasets do not support oxygen limitation of long-term performance. Whether or not oxygen limitations via the metabolic scope, or a lack thereof, have major ecological consequences remains a highly contested question. Here, we propose a general size and trait-based model of energy and oxygen budgets to determine the relative influence of thermal sensitivity of metabolic rates, oxygen limitation, and environmental conditions on thermal performance. We show that oxygen limitation is not necessary to explain performance curves that are correlated with metabolic scope. Oxygen can drastically limit performance and fitness at temperature extremes, but changes in thermal performance are primarily driven by the interplay of changing metabolic rates and species ecology (i.e., food aquisition and mortality). Furthermore, our model reveals that fitness trends with temperature likely oppose trends in growth and efficiency, suggesting a potential explanation for the paradox that species often occur at lower temperatures than their growth-optimum. We suggest that this general model provides a mechanistic underpinning that can provide general and realistic predictions about temperature impacts on ectotherm performance and effects of climate change on species with different metabolic and ecological traits.

@@

There is currently no accepted universal concept or general mechanistic model that explains thermal performance and ecological responses to temperature (\cite{lefevre_are_2016, jutfelt_oxygen-and_2018}). Two main theories are debated as potential general therories describing temeprature influences on ectotherms. The theory of Oxygen and Capacity Limitation of Temperature (OCLTT; \cite{portner_oxygen-and_2010}),  based on the general idea of oxygen limitation for major ecological processes (growth, fitness, etc; see below), provides a conceptual framework rather than a mechanistic theoretical framework, and its major assumptions have been called into question on numerous occasions (\cite[e.g., ][]{lefevre_are_2016, jutfelt_oxygen-and_2018}). The gill-oxygen limitation theory (GOLT) aims to provide a physiological justification for the use of a temperature and oxygen scaled version of the von Bertalanffi growth model as a mechanistic model of thermal performance. The validity of gill limitation of oxygen supply, as well as the ultimate limitation of growth by oxygen in all conditions, is strongly debated (\cite{brander_overconfidence_2013,pauly_sound_2017,lefevre_modelling_2017,lefevre_models_2017}). In addition, the von Bertalanffi growth model only encapsulates a highly abstract view of organism bio-energetics and ecology, with no reference to ecological rates such as food aquisition, mortality and reproduction. As such, it can only represent a very limited set of aspects of species ecology, and cannot on its own account for changes in ecological rates that are often observed with temperature (\cite{englund_temperature_2011,rall_universal_2012}).

To our knowledge, no general theoretical framework exists to quantitatively explain and predict changes in fundamental ecological rates, such as observed changes of attack rates with temperature (\cite{englund_temperature_2011,rall_universal_2012}), from fundamental physiological processes. Instead, and as advocated by the metabolic theory of ecology (\cite{brown_toward_2004}), ecological theory often treats ecological rates as being directly temperature dependant, without a direct link to the underlying physiological drivers (\cite[e.g., ][]{vucic-pestic_warming_2011, guiet_effects_2016}). However, although ecological rates seem to follow some general patterns in the response to temprature (\cite{englund_temperature_2011,rall_universal_2012}), there is also significant heterogeneity between species and trait groups. Since the primary effect of temperature of organisms is on individual physiology, a general model to explain ecological response should be grounded in physiology, and explicitly link physiological constraints and ecology.

@@

This view was encapsulated in the OCLTT (\cite{portner_oxygen-and_2010}), which suggests that the decrease in metabolic scope towards extreme temperatures limits species ability to sustain core functions such as foraging and growth (i.e., functions beyond SM). In some species, however, maximum metabolic rate increases steadily (\cite{lefevre_are_2016, verberk_does_2016}), suggesting that oxygen may not be the limiting factor at high temperatures. Indeed, it has been argued that oxygen is unlikely to determine performance for most species over most of their temperature range as oxygen limits are rarely reached during normal activity (\cite{jutfelt_oxygen-and_2018, holt_climate_2015}).
